# 16. Instrumental variables estimation{-}

## Program 16.1

- Estimating the average causal using the standard IV estimator via the calculation of sample averages
- Data from NHEFS

```{r, results='hide', message=FALSE, warning=FALSE}
library(here)
```

```{r}
#install.packages("readxl") # install package if required
library("readxl")
nhefs <- read_excel(here("data", "NHEFS.xls"))

# some preprocessing of the data
nhefs$cens <- ifelse(is.na(nhefs$wt82), 1, 0)
summary(nhefs$price82)

# for simplicity, ignore subjects with missing outcome or missing instrument
nhefs.iv <- nhefs[which(!is.na(nhefs$wt82) & !is.na(nhefs$price82)),]
nhefs.iv$highprice <- ifelse(nhefs.iv$price82>=1.5, 1, 0)

table(nhefs.iv$highprice, nhefs.iv$qsmk)

t.test(wt82_71 ~ highprice, data=nhefs.iv)
```

## Program 16.2

- Estimating the average causal effect using the standard IV estimator via two-stage-least-squares regression
- Data from NHEFS
```{r}
#install.packages ("sem") # install package if required
library(sem) 

model1 <- tsls(wt82_71 ~ qsmk, ~ highprice, data = nhefs.iv)
summary(model1)
confint(model1)  # note the wide confidence intervals
```

## Program 16.3

- Estimating the average causal using the standard IV estimator via additive marginal structural models
- Data from NHEFS

- G-estimation: Checking one possible value of psi      
- See Chapter 14 for program that checks several values and computes 95% confidence intervals                 

```{r}
nhefs.iv$psi <- 2.396
nhefs.iv$Hpsi <- nhefs.iv$wt82_71-nhefs.iv$psi*nhefs.iv$qsmk

#install.packages("geepack") # install package if required
library("geepack")
g.est <- geeglm(highprice ~ Hpsi, data=nhefs.iv, id=seqn, family=binomial(),
                corstr="independence")
summary(g.est)

beta <- coef(g.est)
SE <- coef(summary(g.est))[,2]
lcl <- beta-qnorm(0.975)*SE 
ucl <- beta+qnorm(0.975)*SE
cbind(beta, lcl, ucl)
```

## Program 16.4

- Estimating the average causal using the standard IV estimator with altnerative proposed instruments
- Data from NHEFS
```{r}
summary(tsls(wt82_71 ~ qsmk, ~ ifelse(price82 >= 1.6, 1, 0), data = nhefs.iv))
summary(tsls(wt82_71 ~ qsmk, ~ ifelse(price82 >= 1.7, 1, 0), data = nhefs.iv))
summary(tsls(wt82_71 ~ qsmk, ~ ifelse(price82 >= 1.8, 1, 0), data = nhefs.iv))
summary(tsls(wt82_71 ~ qsmk, ~ ifelse(price82 >= 1.9, 1, 0), data = nhefs.iv))
```

## Program 16.5

- Estimating the average causal using the standard IV estimator
- Conditional on baseline covariates
- Data from NHEFS
```{r}
model2 <- tsls(wt82_71 ~ qsmk + sex + race + age + smokeintensity + smokeyrs + 
                      as.factor(exercise) + as.factor(active) + wt71,
             ~ highprice + sex + race + age + smokeintensity + smokeyrs + as.factor(exercise) +
               as.factor(active) + wt71, data = nhefs.iv)
summary(model2)
```