# 17. Causal survival analysis: Stata{-}

## Program 17.1

- Nonparametric estimation of survival curves
- Data from NHEFS
```{r, results='hide', message=FALSE, warning=FALSE}
library(here)
```

```{r, out.width="85%", fig.align='center'}
library("readxl")
nhefs <- read_excel(here("data","NHEFS.xls"))

# some preprocessing of the data 
nhefs$survtime <- ifelse(nhefs$death==0, 120, 
                         (nhefs$yrdth-83)*12+nhefs$modth) # yrdth ranges from 83 to 92

table(nhefs$death, nhefs$qsmk)
summary(nhefs[which(nhefs$death==1),]$survtime)

#install.packages("survival")
#install.packages("ggplot2") # for plots
#install.packages("survminer") # for plots
library("survival")
library("ggplot2")
library("survminer")
survdiff(Surv(survtime, death) ~ qsmk, data=nhefs)

fit <- survfit(Surv(survtime, death) ~ qsmk, data=nhefs)
ggsurvplot(fit, data = nhefs, xlab="Months of follow-up",
           ylab="Survival probability",
           main="Product-Limit Survival Estimates", risk.table = TRUE)
```

## Program 17.2

- Parametric estimation of survival curves via hazards model
- Data from NHEFS

```{r, out.width="85%", fig.align='center'}
# creation of person-month data
#install.packages("splitstackshape")
library("splitstackshape")
nhefs.surv <- expandRows(nhefs, "survtime", drop=F) 
nhefs.surv$time <- sequence(rle(nhefs.surv$seqn)$lengths)-1
nhefs.surv$event <- ifelse(nhefs.surv$time==nhefs.surv$survtime-1 & 
                             nhefs.surv$death==1, 1, 0)
nhefs.surv$timesq <- nhefs.surv$time^2

# fit of parametric hazards model
hazards.model <- glm(event==0 ~ qsmk + I(qsmk*time) + I(qsmk*timesq) + 
                       time + timesq, family=binomial(), data=nhefs.surv)
summary(hazards.model)

# creation of dataset with all time points under each treatment level
qsmk0 <- data.frame(cbind(seq(0, 119),0,(seq(0, 119))^2))
qsmk1 <- data.frame(cbind(seq(0, 119),1,(seq(0, 119))^2))

colnames(qsmk0) <- c("time", "qsmk", "timesq")
colnames(qsmk1) <- c("time", "qsmk", "timesq")

# assignment of estimated (1-hazard) to each person-month */
qsmk0$p.noevent0 <- predict(hazards.model, qsmk0, type="response")
qsmk1$p.noevent1 <- predict(hazards.model, qsmk1, type="response")

# computation of survival for each person-month
qsmk0$surv0 <- cumprod(qsmk0$p.noevent0)
qsmk1$surv1 <- cumprod(qsmk1$p.noevent1)

# some data management to plot estimated survival curves
hazards.graph <- merge(qsmk0, qsmk1, by=c("time", "timesq"))
hazards.graph$survdiff <- hazards.graph$surv1-hazards.graph$surv0

# plot
ggplot(hazards.graph, aes(x=time, y=surv)) + 
  geom_line(aes(y = surv0, colour = "0")) + 
  geom_line(aes(y = surv1, colour = "1")) + 
  xlab("Months") + 
  scale_x_continuous(limits = c(0, 120), breaks=seq(0,120,12)) +
  scale_y_continuous(limits=c(0.6, 1), breaks=seq(0.6, 1, 0.2)) +
  ylab("Survival") + 
  ggtitle("Survival from hazards model") + 
  labs(colour="A:") +
  theme_bw() + 
  theme(legend.position="bottom")
```

## Program 17.3

- Estimation of survival curves via IP weighted hazards model
- Data from NHEFS
```{r, out.width="85%", fig.align='center'}
# estimation of denominator of ip weights
p.denom <- glm(qsmk ~ sex + race + age + I(age*age) + as.factor(education)
               + smokeintensity + I(smokeintensity*smokeintensity)
               + smokeyrs + I(smokeyrs*smokeyrs) + as.factor(exercise)
               + as.factor(active) + wt71 + I(wt71*wt71), 
               data=nhefs, family=binomial())
nhefs$pd.qsmk <- predict(p.denom, nhefs, type="response")

# estimation of numerator of ip weights
p.num <- glm(qsmk ~ 1, data=nhefs, family=binomial())
nhefs$pn.qsmk <- predict(p.num, nhefs, type="response")

# computation of estimated weights
nhefs$sw.a <- ifelse(nhefs$qsmk==1, nhefs$pn.qsmk/nhefs$pd.qsmk,
                     (1-nhefs$pn.qsmk)/(1-nhefs$pd.qsmk))
summary(nhefs$sw.a)

# creation of person-month data
nhefs.ipw <- expandRows(nhefs, "survtime", drop=F) 
nhefs.ipw$time <- sequence(rle(nhefs.ipw$seqn)$lengths)-1
nhefs.ipw$event <- ifelse(nhefs.ipw$time==nhefs.ipw$survtime-1 & 
                            nhefs.ipw$death==1, 1, 0)
nhefs.ipw$timesq <- nhefs.ipw$time^2

# fit of weighted hazards model
ipw.model <- glm(event==0 ~ qsmk + I(qsmk*time) + I(qsmk*timesq) + 
                   time + timesq, family=binomial(), weight=sw.a,
                 data=nhefs.ipw)
summary(ipw.model)

# creation of survival curves
ipw.qsmk0 <- data.frame(cbind(seq(0, 119),0,(seq(0, 119))^2))
ipw.qsmk1 <- data.frame(cbind(seq(0, 119),1,(seq(0, 119))^2))

colnames(ipw.qsmk0) <- c("time", "qsmk", "timesq")
colnames(ipw.qsmk1) <- c("time", "qsmk", "timesq")

# assignment of estimated (1-hazard) to each person-month */
ipw.qsmk0$p.noevent0 <- predict(ipw.model, ipw.qsmk0, type="response")
ipw.qsmk1$p.noevent1 <- predict(ipw.model, ipw.qsmk1, type="response")

# computation of survival for each person-month
ipw.qsmk0$surv0 <- cumprod(ipw.qsmk0$p.noevent0)
ipw.qsmk1$surv1 <- cumprod(ipw.qsmk1$p.noevent1)

# some data management to plot estimated survival curves
ipw.graph <- merge(ipw.qsmk0, ipw.qsmk1, by=c("time", "timesq"))
ipw.graph$survdiff <- ipw.graph$surv1-ipw.graph$surv0

# plot
ggplot(ipw.graph, aes(x=time, y=surv)) + 
  geom_line(aes(y = surv0, colour = "0")) + 
  geom_line(aes(y = surv1, colour = "1")) + 
  xlab("Months") + 
  scale_x_continuous(limits = c(0, 120), breaks=seq(0,120,12)) +
  scale_y_continuous(limits=c(0.6, 1), breaks=seq(0.6, 1, 0.2)) +
  ylab("Survival") + 
  ggtitle("Survival from IP weighted hazards model") + 
  labs(colour="A:") +
  theme_bw() + 
  theme(legend.position="bottom")
```

## Program 17.4

- Estimating of survival curves via g-formula
- Data from NHEFS
```{r, out.width="85%", fig.align='center'}
# fit of hazards model with covariates
gf.model <- glm(event==0 ~ qsmk + I(qsmk*time) + I(qsmk*timesq)
                + time + timesq + sex + race + age + I(age*age)
                + as.factor(education) + smokeintensity 
                + I(smokeintensity*smokeintensity) + smkintensity82_71 
                + smokeyrs + I(smokeyrs*smokeyrs) + as.factor(exercise) 
                + as.factor(active) + wt71 + I(wt71*wt71), 
                data=nhefs.surv, family=binomial())
summary(gf.model)

# creation of dataset with all time points for 
# each individual under each treatment level
gf.qsmk0 <- expandRows(nhefs, count=120, count.is.col=F) 
gf.qsmk0$time <- rep(seq(0, 119), nrow(nhefs))
gf.qsmk0$timesq <- gf.qsmk0$time^2
gf.qsmk0$qsmk <- 0

gf.qsmk1 <- gf.qsmk0
gf.qsmk1$qsmk <- 1

gf.qsmk0$p.noevent0 <- predict(gf.model, gf.qsmk0, type="response")
gf.qsmk1$p.noevent1 <- predict(gf.model, gf.qsmk1, type="response")

#install.packages("dplyr")
library("dplyr")
gf.qsmk0.surv <- gf.qsmk0 %>% group_by(seqn) %>% mutate(surv0 = cumprod(p.noevent0))
gf.qsmk1.surv <- gf.qsmk1 %>% group_by(seqn) %>% mutate(surv1 = cumprod(p.noevent1))

gf.surv0 <- aggregate(gf.qsmk0.surv, by=list(gf.qsmk0.surv$time), FUN=mean)[c("qsmk", "time", "surv0")]
gf.surv1 <- aggregate(gf.qsmk1.surv, by=list(gf.qsmk1.surv$time), FUN=mean)[c("qsmk", "time", "surv1")]

gf.graph <- merge(gf.surv0, gf.surv1, by=c("time"))
gf.graph$survdiff <- gf.graph$surv1-gf.graph$surv0

# plot
ggplot(gf.graph, aes(x=time, y=surv)) + 
  geom_line(aes(y = surv0, colour = "0")) + 
  geom_line(aes(y = surv1, colour = "1")) + 
  xlab("Months") + 
  scale_x_continuous(limits = c(0, 120), breaks=seq(0,120,12)) +
  scale_y_continuous(limits=c(0.6, 1), breaks=seq(0.6, 1, 0.2)) +
  ylab("Survival") + 
  ggtitle("Survival from g-formula") + 
  labs(colour="A:") +
  theme_bw() + 
  theme(legend.position="bottom")
```

## Program 17.5

- Estimating of median survival time ratio via a structural nested AFT model
- Data from NHEFS
```{r}
# some preprocessing of the data
nhefs <- read_excel(here("data", "NHEFS.xls"))
nhefs$survtime <- ifelse(nhefs$death==0, NA, (nhefs$yrdth-83)*12+nhefs$modth) # * yrdth ranges from 83 to 92

# model to estimate E[A|L]
modelA <- glm(qsmk ~ sex + race + age + I(age*age)
              + as.factor(education) + smokeintensity
              + I(smokeintensity*smokeintensity) + smokeyrs
              + I(smokeyrs*smokeyrs) + as.factor(exercise)
              + as.factor(active) + wt71 + I(wt71*wt71),
              data=nhefs, family=binomial())

nhefs$p.qsmk <- predict(modelA, nhefs, type="response") 
d <- nhefs[!is.na(nhefs$survtime),] # select only those with observed death time
n <- nrow(d)

# define the estimating function that needs to be minimized
sumeef <- function(psi){
  
  # creation of delta indicator
  if (psi>=0){
    delta <- ifelse(d$qsmk==0 | 
                      (d$qsmk==1 & psi <= log(120/d$survtime)), 
                    1, 0)
  } else if (psi < 0) {
    delta <- ifelse(d$qsmk==1 | 
                      (d$qsmk==0 & psi > log(d$survtime/120)), 1, 0)
  }
  
  smat <- delta*(d$qsmk-d$p.qsmk)
  sval <- sum(smat, na.rm=T)
  save <- sval/n
  smat <- smat - rep(save, n)
  
  # covariance
  sigma <- t(smat) %*% smat
  if (sigma == 0){
    sigma <- 1e-16
  }
  estimeq <- sval*solve(sigma)*t(sval)
  return(estimeq)
}

res <- optimize(sumeef, interval = c(-0.2,0.2))
psi1 <- res$minimum
objfunc <- as.numeric(res$objective)


# Use simple bisection method to find estimates of lower and upper 95% confidence bounds
increm <- 0.1
for_conf <- function(x){
  return(sumeef(x) - 3.84)
}

if (objfunc < 3.84){
  # Find estimate of where sumeef(x) > 3.84
  
  # Lower bound of 95% CI
  psilow <- psi1
  testlow <- objfunc
  countlow <- 0
  while (testlow < 3.84 & countlow < 100){
    psilow <- psilow - increm
    testlow <- sumeef(psilow)
    countlow <- countlow + 1
  }
  
  # Upper bound of 95% CI
  psihigh <- psi1
  testhigh <- objfunc
  counthigh <- 0
  while (testhigh < 3.84 & counthigh < 100){
    psihigh <- psihigh + increm
    testhigh <- sumeef(psihigh)
    counthigh <- counthigh + 1
  }
  
  # Better estimate using bisection method
  if ((testhigh > 3.84) & (testlow > 3.84)){
    
    # Bisection method
    left <- psi1
    fleft <- objfunc - 3.84
    right <- psihigh
    fright <- testhigh - 3.84
    middle <- (left  + right) / 2
    fmiddle <- for_conf(middle)
    count <- 0
    diff <- right - left
    
    while (!(abs(fmiddle) < 0.0001 | diff < 0.0001 | count > 100)){
      test <- fmiddle * fleft
      if (test < 0){
        right <- middle
        fright <- fmiddle
      } else {
        left <- middle
        fleft <- fmiddle
      }
      middle <- (left + right) / 2
      fmiddle <- for_conf(middle)
      count <- count + 1
      diff <- right - left
    }
    
    psi_high <- middle
    objfunc_high <- fmiddle + 3.84
    
    # lower bound of 95% CI
    left <- psilow
    fleft <- testlow - 3.84
    right <- psi1
    fright <- objfunc - 3.84
    middle <- (left + right) / 2
    fmiddle <- for_conf(middle)
    count <- 0
    diff <- right - left
    
    while(!(abs(fmiddle) < 0.0001 | diff < 0.0001 | count > 100)){
      test <- fmiddle * fleft
      if (test < 0){
        right <- middle
        fright <- fmiddle
      } else {
        left <- middle
        fleft <- fmiddle
      }
      middle <- (left + right) / 2
      fmiddle <- for_conf(middle)
      diff <- right - left
      count <- count + 1
    }
    psi_low <- middle
    objfunc_low <- fmiddle + 3.84
    psi <- psi1
  }
}
c(psi, psi_low, psi_high)
```