--- 
title: "Causal Inference: What If. Exercises -- R code"
author: "Book by M. A. Hernán and J. M. Robins, R code by Joy Shi and Sean McGrath, R Markdown code by Tom Palmer"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
#bibliography: [book.bib, packages.bib]
#biblio-style: apalike
link-citations: yes
description: "Code examples from the Causal Inference Book by M. A. Hernán and J. M. Robins https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/"
geometry: margin=1in
fontsize: 10pt
linestretch: 1.1
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = 'formatR', digits = 4, tidy.opts = list(width.cutoff = 60)) 
```

# Preface{-}

This book presents code examples from @ci-book, which is available in draft form from the following webpage.

https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/

The R code is based on the code by Joy Shi and Sean McGrath given [here](https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/1268/20/Rcode_CIpart2.zip).

## Packages to install
To install the R packages required for this book please copy/fork the repository and run:
```{r eval=FALSE}
# install.packages("devtools") # uncomment if devtools not installed
devtools::install_deps()
```

## Downloading the datasets
We assume that you have downloaded the data from the Causal Inference Book website and saved it to a `data` subdirectory. You can do this manually or with the following code (nb. we use the [`here`](https://here.r-lib.org/) package to reference the data subdirectory).

```{r, results='hide', message=FALSE, warning=FALSE}
library(here)
```

```{r}
dataurls <- list()
dataurls[[1]] <- 
  "https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/2012/10/nhefs_sas.zip"
dataurls[[2]] <- 
  "https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/2012/10/nhefs_stata.zip"
dataurls[[3]] <- 
  "https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/2017/01/nhefs_excel.zip"
dataurls[[4]] <- 
  "https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/1268/20/nhefs.csv"

temp <- tempfile()
for (i in 1:3) {
  download.file(dataurls[[i]], temp)
  unzip(temp, exdir = "data")
}

download.file(dataurls[[4]], here("data", "nhefs.csv"))
```

\mainmatter
